---
title: "ngsReports::FastQC"
subtitle: "`r getwd()`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    toc: yes
params:
    dataType: "Transcriptome"
    species: "Hsapiens"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
											results = "asis",
											message = FALSE, warning = FALSE,
											error = FALSE)
```


# Setup

```{r}
library(ngsReports)
library(dplyr)
library(ggplot2)
library(tibble)
library(pander)
library(scales)
library(magrittr)
library(stringr)
library(readr)
```

```{r setGlobals}
globals <- list(usePlotly = TRUE,
								cluster = TRUE,
								dendrogram = TRUE,
								theoreticalGC = TRUE,
								theoreticalType = params$dataType,
								species = params$species)
```


```{r rawFastqcData}
rawFastqcFiles <- list.files(pattern = "(fastqc.zip|fastqc)$")
stopifnot(length(rawFastqcFiles) > 1)
message("FastQC files found. Loading FastQC data")
rawFastqcData <- tryCatch(getFastqcData(rawFastqcFiles))
plotLabels <- structure(gsub(".(fastq|fastq.gz|bam)", "", fileName(rawFastqcData)),
												names = fileName(rawFastqcData))
```


```{r}
message("Checking for targets.csv")
tgtsFile <- "targets.csv"
if (file.exists(tgtsFile)){
	message("Found targets.csv...checking columns")
	targets <- read_csv(tgtsFile)
	fCol <- grep("[Ff]ile[Nn]ame", colnames(targets))
	lCol <- grep("[Ll]abel", colnames(targets))
	if (length(fCol) == 1 && length(lCol) == 1){
		stopifnot(all(fileName(rawFastqcData) %in% targets[[fCol]]))
		message("Alternate labels found")
		plotLabels <- structure(targets[[lCol]], names = targets[[fCol]])
	}
	else{
		message("No valid labels found")
	}
}
if (!file.exists(tgtsFile)){
	message("targets.csv not found. Using default labels")
}
```


# FastQC Summary


```{r}
Basic_Statistics(rawFastqcData) %>%
	mutate(Sequence_length = paste(Shortest_sequence, Longest_sequence, sep = "-")) %>%
	dplyr::select(Filename, contains("sequence"), `%GC`, -contains("est")) %>%
	set_names(gsub("_", " ", names(.))) %>%
	set_names(str_to_title(names(.))) %>%
	set_names(gsub("Sequences", "", names(.))) %>%
	set_names(str_trim(names(.))) %>%
	rename(`%GC` = `%Gc`) %>%
	pander(big.mark = ",",
				 split.tables = Inf,
				 style = "rmarkdown",
				 justify = "lrrrr",
				 caption = "Summary statistics for all libraries")
```

## Read Totals

Library Sizes ranged between `r pander(comma(range(readTotals(rawFastqcData)$Total_Sequences)))` reads.

```{r plotReadTotals, fig.cap = "Read totals for each library. Duplicated reads are conventionally an high overestimate at this point."}
plotReadTotals(rawFastqcData, labels = plotLabels, usePlotly = globals$usePlotly)
```


## FastQC Summary

```{r plotSummary, fig.cap="Summary of FastQC flags for each parameter"}
plotSummary(rawFastqcData, labels = plotLabels, usePlotly = globals$usePlotly)
```

## Per Base Sequence Quality

```{r plotBaseQualities, fig.cap = "Heatmap showing mean base qualities for each library"}
plotBaseQualities(rawFastqcData, labels = plotLabels, cluster = globals$cluster, dendrogram = globals$dendrogram, usePlotly = globals$usePlotly)
```

## Per Sequence Quality Scores

```{r plotSequenceQualitiesHeatmap, fig.cap = "Heatmap showing mean sequence qualities for each library"}
plotSequenceQualities(rawFastqcData, labels = plotLabels, cluster = globals$cluster, dendrogram = globals$dendrogram, usePlotly = globals$usePlotly)
```

## Per Base Sequence Content

```{r plotSequenceContent, fig.cap = "Heatmap of summed base distributions along each read"}
plotSequenceContent(rawFastqcData, labels = plotLabels, cluster = globals$cluster, dendrogram = globals$dendrogram, usePlotly = globals$usePlotly)
```

## Per Sequence GC Content

```{r plotGCHeat, fig.cap = paste("GC Content Heatmap normalised to theoretical GC content in the", globals$species, globals$theoreticalType)}
plotGcContent(rawFastqcData, labels = plotLabels, theoreticalGC = globals$theoreticalGC, theoreticalType = globals$theoreticalType, species = globals$species, cluster = globals$cluster, dendrogram = globals$dendrogram, usePlotly = globals$usePlotly)
```


```{r plotGCLine, fig.cap = paste("GC Content Distributions for all reads showing theoretical GC content from the", globals$species, globals$theoreticalType)}
plotGcContent(rawFastqcData, labels = plotLabels, theoreticalGC = globals$theoreticalGC, theoreticalType = globals$theoreticalType, species = globals$species, plotType = "line", usePlotly = globals$usePlotly)
```

## Sequence Length Distribution

```{r plotSequenceLengthDistribution}
plotSequenceLengthDistribution(rawFastqcData, labels = plotLabels, cluster = globals$cluster, dendrogram = globals$dendrogram, usePlotly = globals$usePlotly)
```

## Sequence Duplication Levels

```{r plotDuplicationLevels}
plotDuplicationLevels(rawFastqcData, labels = plotLabels, cluster = globals$cluster, dendrogram = globals$dendrogram, usePlotly = globals$usePlotly)
```

## Overrepresented Sequences

```{r Overrepresented_sequences, echo=FALSE, results='asis', message=TRUE}
nOver <- 30
os <- Overrepresented_sequences(rawFastqcData) 
if (nrow(os) > 0){
	os %>% 
		mutate(Filename = plotLabels[Filename]) %>%
		arrange(Sequence) %>%
		dplyr::slice(1:nOver) %>%
		pander(split.tables = Inf,
					 style = "rmarkdown",
					 justify = "llrrl",
					 big.mark = ",",
					 caption = paste("Summary of Overrepresented Sequences in all files.",
					 								"A maximum of", nOver, "seqences are shown."))
}
if (nrow(os) == 0){
	message("No overrepresented sequences were detected by FastQC")
}
```

## Adapter Content

```{r plotAdapterContent, fig.cap="Universal Adapter Content"}
plotAdapterContent(rawFastqcData, labels = plotLabels, cluster = globals$cluster, dendrogram = globals$dendrogram, usePlotly = globals$usePlotly) 
```

## Kmer Content

```{r plotKmers}
plotKmers(rawFastqcData, labels = plotLabels, usePlotly = globals$usePlotly, cluster = globals$cluster, dendrogram = globals$dendrogram)
```

```{r Kmer_Content, message=TRUE}
nK <- 30
kc <- Kmer_Content(rawFastqcData)
if (nrow(kc) > 0){
  kc %>%
    arrange(desc(Count)) %>%
    dplyr::slice(1:nK) %>%
    arrange(Sequence) %>%
    mutate(Filename = plotLabels[Filename]) %>%
    pander(split.tables = Inf,
           style = "rmarkdown",
           justify = "llrrrr",
           big.mark = ",",
           caption = paste("Summary of Overrepresented Kmers in all files.",
                           "A maximum of", nK, "Kmers are shown."))
}
if (nrow(kc) == 0){
  message("No Kmers were detected by FastQC")
}
```


